---
title: "Building CDMetaPOP input files with the R wrapper"
subtitle: "Create ClassVars, PatchVars, PopVars and RunVars objects, and launch a simulation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building CDMetaPOP input files in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = "#>"
)

library(cdmetapopR)
```

# Overview

CDMetaPOP input files consist of a series of connected csv files: a `RunVars.csv` naming one
or more `PopVars.csv` files, each naming a `PatchVars.csv`, which in turn names
a `ClassVars.csv` plus cost-distance matrices, gene frequency files, and other inputs. This wrapper
is mean to help the user create and assemble these files and handle the directories and relative 
paths of the different input files under the hood. 

`cdmetapopR` represents these file types as R objects based on unique classes:

| Object | File | Each row is |
|---|---|---|
| `ClassVars()` | `ClassVars.csv` | an age class |
| `PatchVars()` | `PatchVars.csv` | a patch |
| `PopVars()` | `PopVars.csv` | a batch |
| `RunVars()` | `RunVars.csv` | a run |

Default values for each input class match values in CDMetaPOP's own example files,
each column can be modified by the user, after which the wrapper will validate
everything assigned to an input object, and can hold another object wherever the inputs call
for another object. 'launch_cdmetapop()' will write the entire input file directory for you, 
and you never have to name a csv file or manage subdirectories. If your CDMetaPOP input files 
already exist, you can read them in as R objects using 'read_cdmetapop()' or using the
'path' argument in each of the above functions.

Here is on example that will create all necessary inputs:

```{r overview-sketch, eval=FALSE}
mycv     <- ClassVars()
mypv     <- PatchVars(class_vars = mycv)       # an object, not a path
mypopv   <- PopVars(xyfilename = mypv)
myrv     <- RunVars(Popvars = mypopv)
run_dir  <- launch_cdmetapop(myrv, pythonFilepath = "python",
                             CDMetaPOPFilepath = "CDMetaPOP.py")
```

Even simpler, you can run the default example simulation by just calling 'RunVars()' 
in the 'launch_cdmetapop()' function.

```{r launch-example, eval=FALSE}
run_dir  <- launch_cdmetapop(RunVars(), pythonFilepath = "python",
                             CDMetaPOPFilepath = "CDMetaPOP.py")
```

`write_cdmetapop()` can be used to write a single object to file, `read_cdmetapop()` or the `path =`
argument can be used to read an existing file in as an R object.

This vignette builds a complete, valid input set from scratch. Everything except
the launch itself runs as shown.

# ClassVars

`ClassVars.csv` holds the age-class parameters: size at age, mortality,
maturation, migration probability, etc. A default call reproduces
CDMetaPOP's `ClassVars_AS1.csv`, which has six age classes.

```{r classvars-default}
mycv <- ClassVars()
mycv$ages
```

Columns are read and written as vectors with one value per age class. Printing
the object shows the table; `as.data.frame()` gives you an ordinary data frame
to inspect or subset.

```{r classvars-inspect}
as.data.frame(mycv)[, c("Age class", "Body Size Mean", "Age Mortality Out")]
```

Deafult columns can be modified when the function is called, or already created 
objects can be modified just like a data frame. Input values
are validated immediately, so incorrect input values will be flagged immediately:

```{r classvars-modify}
mycv$age_mortality_out <- c(0.6, 0.4, 0.2, 0.1, 0.1, 0.1)
mycv$maturation <- c(0, 0, 0.5, 1, 1, 1)
mycv$age_mortality_out

# Alternatively, edit the columns at creation
mycv <- ClassVars(age_mortality_out = c(0.6, 0.4, 0.2, 0.1, 0.1, 0.1),
                  maturation = c(0, 0, 0.5, 1, 1, 1))
```

`add_rows()` adds age classes, copying the current last row so you only edit
what differs:

```{r classvars-addrow}
add_rows(mycv)
mycv$ages
mycv$maturation
```

# PatchVars

`PatchVars.csv` contains one row for each patch and includes spatial coordinates, carrying capacity,
initial population size, mortality, migration, and growth parameters. The default is
CDMetaPOP's seven-patch `PatchVarsS1.csv`.

The `class_vars` column is where a `PatchVars` file would normally name a
`ClassVars.csv`. With the wrapper, you can provide either the R object
or the relative filepath:

```{r patchvars-build}
mypv <- PatchVars(class_vars = mycv, genes_initialize = "random")
mypv$patch_id
```

`genes_initialize = "random"` draws random starting genotypes rather than reading an
allele-frequency file.

Edit columns one value per patch, and add patches with `add_rows()`:

```{r patchvars-modify}
mypv$k <- c(300, 300, 500, 500, 200, 200, 200)
mypv$n0 <- 100

# Alternatively,
mypv <- PatchVars(k = c(300, 300, 500, 500, 200, 200, 200),
                  n0 = 100)

add_rows(mypv)
mypv$patch_id
mypv$k
```

# Changing parameter values over time and by sex

CDMetaPOP packs multiple values into one cell using delimiters, be careful to 
get the formatting correct.

| Delimiter | Meaning | Example |
|---|---|---|
| &#124; | changes over time, at the years set by `RunVars$cdclimgentime` | `"0|0.2|0.5"` |
| `~` | differs by sex, in the order FXX, MXY, MYY, FYY | `"0.5~0.5"` |
| `;` | multiple species, or multiple files for one field | `"popvars/speciesA.csv;popvars/speciesB.csv"` |

**A delimited value is a single string, not a vector.** 

```{r delimiters}
# One patch's mortality changing over time: 0, then 0.2, then 0.5
mypv$mortality_out <- "0|0.2|0.5"

# Eight patches, each with different moratlity values over time
mypv$mortality_out <- c("0|0.2|0.2","0|0.2|0.3","0|0.2|0.4","0|0.2|0.5","0|0.2|0.6","0|0.2|0.7","0|0.2|0.8","0|0.2|0.8")

# Eight patches, each with its own constant mortality
mypv$mortality_out <- c(0, 0.1, 0.1, 0.2, 0.2, 0.2, 0.3, 0.3)
```

The first assigns the same three-stage schedule to every patch. The second gives
each patch different values over time, and the third gives one value for each patch. 
Writing `c(0, 0.2, 0.5)` when you meant `"0|0.2|0.5"` won't raise any errors but 
is not correct and won't be caught.

Each segment is validated independently, so an out-of-range value inside a
delimited string is still rejected:

```{r delimiters-validated, error=TRUE}
mypv$mortality_out <- "0|0.2|1.5"
```

```{r delimiters-reset, include=FALSE}
mypv$mortality_out <- c(0, 0.1, 0.1, 0.2, 0.2, 0.2, 0.3, 0.3)
```

Sex-specific values work the same way, as one `~`-delimited string:

```{r delimiters-tilde}
# XX~XY
mycv$sex_ratio <- "0.5~0.5"
```

# Cost-distance matrices

`PopVars` has five movement fields (mating, emigration, immigration, straying
and dispersal) that CDMetaPOP reads as patch-by-patch cost-distance
matrices. Each accepts either a path to a csv or **a raw R matrix** which will
be created and serialized for you at launch.

`create_cdmat()` builds a distance matrix from patch coordinates:

```{r cdmat}
coords <- as.data.frame(mypv)[, c("X", "Y")]
cdmat <- create_cdmat(coords, method = "euclidean")

dim(cdmat)
round(cdmat[1:4, 1:4])
```

`method = "equal"` gives every patch pair the same cost, and `method = "lcp"`
computes least-cost paths across a resistance raster. See the main vignette for
more on `create_cdmat()`.

# PopVars

`PopVars.csv` holds one row per *batch* - a set of population-level settings
that CDMetaPOP runs as an independent simulation. The `xyfilename` column points at a
`PatchVars` file, so hand it the object, and give the five movement fields the matrix
built above:

```{r popvars-build}
mypopv <- PopVars(
  n_batches = 1,
  xyfilename = mypv,
  mate_cdmat = cdmat,
  migrateout_cdmat = cdmat,
  migrateback_cdmat = cdmat,
  stray_cdmat = cdmat,
  disperseLocal_cdmat = cdmat,
  correlation_matrix = "N",
  subpopmort_file = "N",
  betaFile_selection = "N"
)
```

`"N"` is CDMetaPOP's off switch, used here for the three optional fields so that
nothing outside this session is referenced.

`PopVars` has 81 columns, and unlike `ClassVars`/`PatchVars` CDMetaPOP reads its
headers as dictionary keys so the `PopVars()` argument and the csv column names 
match exactly:

```{r popvars-modify}
mypopv$loci <- 4
mypopv$matemoveno <- "4"

add_rows(mypopv)
mypopv$popmodel <- c("packing", "logistic_back")
as.data.frame(mypopv)[, c("loci", "alleles", "matemoveno", "popmodel")]
```

The second batch reuses the same `PatchVars` object with a different set
of population parameters, which CDMetaPOP will run as a separate simulation.

# RunVars

`RunVars.csv` is the top of the file tree. Each row is a *run*, and its `Popvars`
column names one `PopVars` file per species. CDMetaPOP executes every run × every
batch × every Monte Carlo replicate as a separate simulation.

```{r runvars-build}
myrv <- RunVars(n_runs = 1, Popvars = mypopv)
myrv$runtime <- 50
myrv$output_years <- 5
myrv
```

The set of R input objects is now complete: `myrv` → `mypopv` → `mypv` → `mycv`, plus the
cost matrix. Nothing has been written to disk at this point, and no file has been named.

# Writing and reading single files

The normal route writes everything at launch (next section). But when you want
one file on its own to inspect, hand-edit, send to a collaborator, or
feed it to one of the `make_*()` Shiny apps, `write_cdmetapop()` writes a single
object to a csv you name:

```{r write-one}
out_dir <- file.path(tempdir(), "cdmetapop_inputs")
dir.create(out_dir, showWarnings = FALSE)

write_cdmetapop(mycv, file.path(out_dir, "ClassVars.csv"))
```

Read one back either with `read_cdmetapop()`, or by passing `path =` to the
constructor. The two are equivalent:

```{r read-two-ways}
cv_a <- read_cdmetapop(file.path(out_dir, "ClassVars.csv"), type = "ClassVars")
cv_b <- ClassVars(path = file.path(out_dir, "ClassVars.csv"))

identical(as.data.frame(cv_a), as.data.frame(cv_b))
identical(as.data.frame(cv_a), as.data.frame(mycv))
```

`path =` has the advantage that you can override columns as you read:

```{r read-override}
cv_c <- ClassVars(path = file.path(out_dir, "ClassVars.csv"), maturation = 1)
cv_c$maturation
```

The same works for all four classes. 

There is one important limit. `write_cdmetapop()` deliberately does **not**
follow references: a cell holding a path is written verbatim, but a cell holding
a live object or a raw matrix has no path to write, so it is written as `NA` 
with a warning:

```{r write-warning, warning=TRUE}
write_cdmetapop(myrv, file.path(out_dir, "RunVars_nested.csv"))
```

A file written that way cannot be read back in immediately or launched with `launch_cdmetapop`,
since `NA` is not a valid value. 
So `write_cdmetapop()` is more for getting an object onto disk to look at,
 hand-edit, or prep for a direct Python launch.

# Launching a simulation

`launch_cdmetapop()` takes the `RunVars` object, writes every file in the set
into a fresh timestamped run directory, and calls Python to start CDMetaPOP:

```{r launch, eval=FALSE}
run_dir <- launch_cdmetapop(
  myrv,
  pythonFilepath = "C:/Users/User1/anaconda3/python.exe",
  CDMetaPOPFilepath = "C:/Users/User1/CDMetaPOP/src/CDMetaPOP.py"
)
```

Runs can take days, so the simulation launches in the background by default and
the function returns immediately. Pass `wait = TRUE` to block until it finishes.
On macOS and Linux a background run opens no console window, so check
CDMetaPOP's `CDmetaPOP0.log` in the output folder for progress and errors.

You can also point it at a directory of existing csvs instead of an object:

```{r launch-path, eval=FALSE}
launch_cdmetapop("C:/Users/User1/CDMetaPOP/example_files", wait = TRUE)
```

## What gets written

The run directory holds CDMetaPOP's canonical layout, with every cross-reference
rewritten to the right relative path:

```
cdmetapop_run_20261002_143015/
├── RunVars.csv
├── popvars/mypopv.csv
├── patchvars/mypv.csv
├── classvars/mycv.csv
├── cdmats/cdmat.csv
└── manifest.csv
```

Files are named after the R variables that held the objects, so you can trace
any csv back to where it came from. `manifest.csv` records that mapping
explicitly:

# Validation

Values are checked as you assign them, against the rules in CDMetaPOP's user
manual, so problems surface at the point you introduce them:

```{r validation-enum, error=TRUE}
myrv$gridformat <- "csv"
```

```{r validation-range, error=TRUE}
mypv$straying_prob <- 1.5
```

```{r validation-length, error=TRUE}
mycv$maturation <- c(0, 1)
```

Cross-file consistency is checked at launch, where the whole file set is visible at
once. For example, a `PopVars` whose referenced `PatchVars` is missing something
it needs. Those errors are raised before any directory is created, so a failed
launch leaves nothing behind to clean up.

# Reading results

`launch_cdmetapop()` returns the run directory, which is what the `summary_*()`
functions take. Once a run has finished:

```{r read-results, eval=FALSE}
summary_pop(run_dir, type = "N_Initial")
```

Using the example output bundled with the package:

```{r read-results-live, fig.width=7, fig.height=4.5, out.width="100%"}
example_out <- system.file("extdata", "Example_dat", package = "cdmetapopR")
summary_pop(example_out, type = "N_Initial")
```

See the main vignette, *Reading, summarizing, customizing, and plotting
CDMetaPOP outputs*, for the full set of summary and plotting functions.

```{r cleanup, include=FALSE}
unlink(out_dir, recursive = TRUE)
```
